Inflammation, SARS-CoV-2 infection and mental health disorders impact on systemic serotonin and dopamine metabolism

### 2023-05-23 paper/markdown/rebuttal_letter.Rmd

<span custom-style = 'reviewer'>__Introduction:__ The use of subheadings would be helpful to facilitate the understanding of the first part of the introduction.</span>

This point is no relevant anymore, since we re-wrote and consolidated __Introduction__ of the revised manuscript. 
We describe each thread in a separate paragraph.

## Issue 15

<span custom-style = 'reviewer'>__Results:__ The statement that TRP is a precursor of IDO is not correct!</span>

Thanks for careful lecture, this erroneous statement was removed.

## Issue 16

<span custom-style = 'reviewer'>__Results:__ The description of the time between the infection and the assessment of 138.5 days leads to the question of how the time between the infection and the assessment was assessed. Maybe it would be better to use complete instead of half and quarter days as units.</span>

This time interval refereed to the SARS-CoV-2 infection diagnosis via PCR test, which we specify in the revised manuscript. 
We also round them to complete days, as suggested.

## Issue 17

<span custom-style = 'reviewer'>__Results:__ h. Table 1: I would suggest adding further lines in the table as "heading" lines specifying the cohorts described in the following parts.</span>

For the sake of clarity, we present characteristic of the SIMMUN and INCOV cohorts in separate tables (__Table 1__ and __Table 2__).

## Issue 18

<span custom-style = 'reviewer'>__Results:__ Page 9 lines 24-41; This paragraph should be in the Methods instead of the Results. Furthermore, the rationale is missing: Why was INCOV used as a comparison or additional study to your own data? In addition, information on the measurement /measurement system (for INCOV) should be provided. Were the measuring probes used in SIMMUN and INCOV comparable?</span>

We concur with your suggestion, the variables of interests are now listed in the __Methods__ section instead of __Results__. 
Furthermore, we describe the rationale for analyses in the INCOV cohort in the last paragraph of __Introduction__, __Methods/Analysis endpoints__ and __Results__. 
In brief, we employed the SIMMUN collective providing more extensive demographic and clinical characteristic in an initial search for the most relevant factors impacting on the indirect readouts of serotonin and dopamine metabolism: TRP, KYN, PHE and TYR [@Brown2021; @Marx2020; @Capuron2011; @Widner1997; @Neurauter2008]. 
The INCOV dataset with measurements of more direct markers of the neurotransmitter availability, i.e. plasma serotonin and dopamine 3-O-sulfate, the most abundant dopamine metabolite present in human blood [@Meiser2013; @Goldstein1999], allowed us to validate effects of inflammation, SARS-CoV-2 infection and age, i.e. significant predictors identified in the SIMMUN study. 
Furthermore, due to longitudinal design of the INCOV study, we were able to explore not only the impact of infection recovery but also acute and sub-acute COVID-19 on serotonin and dopamine turnover. 
Still, the incompatibility of designs and datasets of these two collectives hindered us from developing consistent models in one cohort and direct validation in the outer (canonical training - test or external validation setup). 
We list this issue in __Limitations__.

## Issue 19

<span custom-style = 'reviewer'>__Results__: The term acute and sub-acute should be defined.</span>

We provide the time intervals for the acute, sub-acute and recovery timepoints in the INCOV cohort in the revised manuscript (__Methods__, __Results__, __Figures__). 
Of note, those time intervals fit well with the definition of acute, sub-acute and long-COVID/recovery timepoints in our recent population study [@Sahanic2021].

## Issue 20

<span custom-style = 'reviewer'>__Results:__ What does the term "virtually age-independent" mean? I would appreciate it if concrete numbers are stated here instead.</span>

We apologize for this imprecise term, it was removed from the revised manuscript.

## Issue 21

<span custom-style = 'reviewer'>__Results:__ Page 11 lines 35 to 54 should be part of the statistical description in the Method section.</span>

This part of the manuscript was removed during revision. 
Following your suggestion, we keep details on procedures and statistical analysis in the __Methods__ section.

## Issue 22

<span custom-style = 'reviewer'>__Results:__ Page 15 line 28: standard inflammatory cytokines should be named more specifically.</span>

We are sorry for this unclarity. 
In the revised text, we name inflammatory markers (NEO, NLR, cytokines) more explicitly.

## Issue 23

<span custom-style = 'reviewer'>__Discussion:__  The discussion of the study results is rather shallow: Results should be better integrated into existing evidence. Furthermore, there is no discussion of the results of depressive symptoms (HADS).</span>

In the revised manuscript, we restructured and improved scientific quality of __Discussion__. 
The key explanatory factors proposed by us to influence systemic serotonin and dopamine availability: age, inflammation, SARS-CoV-2 infection and, for serotonin, availability of the precursor aminoacid TRP, are discussed in a context of literature evidence. 
We also critically elaborate on the link between plasma serotonin, TRP and KYN levels and depressive disorders, which was recently challenged [@Capuron2011; @Hufner2021; @Sun2020; @Hufner2015; @Fellendorf2022; @Paul-Savoie2011; @Hunt2020; @Moncrieff2022; @Marx2020; @Brown2021].

## Issue 24

<span custom-style = 'reviewer'>__Discussion:__ The statement that "BH4 is also important for nitric oxide synthesis and therefore involved in oxidative stress" is not correct. (Page 16) BH4 is an essential Co-factor of NO-synthases that produce NO with its function as a vasodilator and additional anti-pathogenic effects. Only if BH4 is missing, there is an uncoupling of NOS with subsequently increased reactive oxygen/nitrogen species. Thus, your "therefore involved in oxidative stress" is misleading in normal physiological states.</span>

Following discussion in the study team, we decided to suppress the ROS thread of the discussion since it is likely not so important for interpretation of our data. 
Instead we focus on direct effects of inflammation on PAH, the enzyme catalyzing the PHE - TYR conversion, i.e. the initial step of dopamine synthesis and its crucial co-factor BH4 [@Geisler2013; @Luporini2021; @Fanet2021; @Capuron2011; @Neurauter2008].

## Issue 25

<span custom-style = 'reviewer'>__Discussion:__ Page 14 line 60: "pre-pandemic inflammation markers" is probably not the right term. Maybe rewrite the sentence.</span>

We apologize for this obscure term. 
We meant levels of inflammatory markers in SARS-CoV-2-uninfected individuals, which we state now explicitly.

## Issue 26

<span custom-style = 'reviewer'>__Highlights:__ In the Highlights, the results and perhaps the conclusion should be emphasized more than the theoretical background (Highlights 1 and 2) or the methodology (Highlights 3 and 4).</span>

_Katharina: please respond, I don't have this manuscript part at all in pre-revision manuscript files._

## Issue 28

<span custom-style = 'reviewer'>__General:__ Furthermore, there are several grammatical flaws and spelling mistakes.
Some of these mistakes are listed here. 
However, the manuscript would benefit from professional language correction.
</span>

We are grateful for the careful language check. 
We have paid particular attention to the style, grammar and proper scientific language in the revised manuscript. 
For particular language issues:

* _Page 3 line 24: mediate instead of mediated._ - corrected as suggested

* _Page 3 line 29: sentence incomplete_ - rewritten

* _Page 4 l. 35. Results of the SIMMUN study_ - does not apply, __Results__ were rewritten

* _Page 4: line 37: in-person study visit sounds not familiar: better use on-site investigation_ - we use the wording 'on-site study visit' in the revised manuscript

* _Page 4 line 38 is grammatically not correct: probably better […] the results of which are re-ported here_ - rewritten

* _Page 5 line 14: The term physician assessment should be replaced by a better term_ - we use 'general medical assessment' instead.

* _Page 6 line 11: The hyphen is missing at "inflammatory"_ - does not apply, the section was rewritten

* _Page 8: lines 55-56: In the participants who were tested…_ - rewritten

* _Page 8 line 58: timepoint in [the?] the course sounds strange (and the article is missing), may-be better time between infection and assessment._ - changed to 'timepoints of SARS-CoV-2 infection' and 'days after diagnosis via PCR'

* _Page 9 line 52: What are nadir TRP concentrations?_ - we do not use the term any more, changed to 'minimum'

* _Page 11 line 7: a comma is missing._ - does not apply, the section was rewritten.

* _Page 12 line 24: "on the other hand" without "on the one hand"_ - we removed this colloquial phrase and other similar wording from the revised text

* _Page 13: line 4: a comma is missing_ - rewritten

* _Page 14 line 14: […] via their bidirectional influence_ - rewritten

* _Page 14 line 34: please delete the term "larger"_ - done as suggested.

* _Page 15, line 35: of in individuals_ - rewritten

* _Page 16, line 16: incomplete_ - rewritten

## Issue 29

<span custom-style = 'reviewer'>__Supplementary Material:__ As a criterion to compare models, the Bayesian information criterion is stated with AIC as an abbreviation.</span>

We apologize for the typo, it was corrected to 'BIC'.

## Issue 30

<span custom-style = 'reviewer'>__Supplementary Figure S2:__ It is not clear why fitted generalized adaptive models are presented as graphs, as the associated statistic (Spearman's rho) assesses monotone relationships.</span>

We concur with the Reviewer's view, that visualization of trends for monotone associations is problematic. 
Due to profound changes in the manuscript structure, we do not show scatter plots to illustrate monotone correlations between non-normally distributed variables. 
Correlation analysis results in the INCOV cohort are presented now as simple bubble correlograms or visualized as force-directed graphs [@Briatte2021; @Csardi2006].

\newpage

# Reviewer 2

## Issue 31

<span custom-style = 'reviewer'>In this paper, the authors show the relationship between alterations in markers of inflammation, aminoacid precursors, and symptoms of depression, anxiety, and stress in individuals COVID-19-positive or not. 
They relate the results of these observations to the results obtained in another immune and metabolic study carried out at different time points of the COVID-19 trajectory and already published by a group of different investigators.
The methodology is excellent as can be expected from a research team specialized in the study of inflammation and its impact on the kynurenine metabolism pathway. 
The results are original and sufficiently important to be published. 
The main problem with the current manuscript is the confusion coming from the mode of presentation that mixes up in the results section data from the two studies. 
The discussion is circumvoluted and difficult to follow.</span>

We thank the Reviewer for appreciation of out text and the constructive feedback. 
Weaknesses in structure of the test were already pointed out be the Reviewer 1. 
We addressed them by employing the multi-parameter modeling results for identification of the most vital explanatory factors that affect neurotransmitter metabolism and subsequent more specific correlation analyses and statistical testing with such relevant predictors. 
To tackle with the confusion around the two investigated cohorts, we separated the results obtained with the SIMMUN and INCOV collective. 
In more details, we employ the SIMMUN cohort as a kind of 'exploratory' collective to identify factors affecting TRP, KYN, KYN/TRP, PHE and PHE/TYR among a wider palette of demographic, clinical, psychometric, inflammatory and SARS-CoV-2 associated variables. 
Subsequently, we attempted to validate the most important of them: age, inflammation and SARS-CoV-2 infection in the INCOV cohort. 
Finally, we re-wrote the __Discussion__ section with a special focus on the crosstalk between inflammation and neurotransmitter biosynthesis and, in part controversial, relevance of systemic neurotransmitter availability for the central nervous system and mental health.

## Issue 32

<span custom-style = 'reviewer'>There are a few typos in the text, e.g., othe (line 6 instead of other).</span>

Thanks for the careful lecture. 
We paid special attention to the grammar and spelling of the revised text.

## Issue 33

<span custom-style = 'reviewer'>The text would benefit from further editing by a specialized scientific English editing service</span>

_Katharina: could you provide an answer?_

## Issue 34

<span custom-style = 'reviewer'>Line 7: by collectives you probably mean cohorts or population samples.</span>

The term 'collective' has the same meaning as 'cohort' in the manuscript text. 
For practical reasons, we avoid the wording 'population sample', which may suggest a random and hence representative draw from the general population. 
In fact neither the SIMMUN cohort enriched in psychiatric patients not the INCOV cohort enriched in hospitalized COVID-19 patients pose a representative sample of the general population during the pandemic (see __Limitations__) .

## Issue 35

<span custom-style = 'reviewer'>The introduction ends with the statement that the study was conducted in an exploratory manner. Was that the objective of the present study in the absence of any specific hypothesis?</span>

As pointed out by Reviewer 1 as well, the last paragraph of __Introduction__ was quite obscure. 
In the final part of __Introduction__, we present now the 'evidence gap', which is the still incompletely characterized effect of SARS-CoV-2 infection on neurotransmitter metabolism, as well as the aim of our analysis, which identification of factors affecting the neurotransmitter turnover in a cohort of SARS-CoV-2-infected individuals. 
Still, our study does not have a fully exploratory character, since the candidate explanatory factors: age, sex, body mass class, mental and somatic conditions, smoking and alcohol, depression, anxiety, stress and inflammation have already been described to affect serotonin and dopamine metabolism.

## Issue 36

<span custom-style = 'reviewer'>Page 9, line 34 and elsewhere in the text when referring to immune factors that are measured, use the term variables instead of parameters.</span>

We agree that the term 'parameter' may be misleading, especially in context of modeling and its results. 
We changed it to 'variable' throughout the revised text.

## Issue 37

<span custom-style = 'reviewer'>Legend of Figures: the use of the term decay is inappropriate, you probably mean metabolites</span>

We avoid using the term 'decay' in the revised text and utilize 'catabolism' or 'conversion' instead.

## Issue 38

<span custom-style = 'reviewer'>The results are difficult to read as you mix up the two cohorts both in the text and in the figures. It would be preferable to first present the results of your own cohort, the SIMMUN cohort, and then test the generality of the results by assessing whether the same variations are found in the INCOV cohort with the limitation that this can only be done for the relationship between immune factors and neurotransmitter precursors.</span>

We concur with your opinion and present the analysis results as suggested. 
Please refer to __Issues 1__ and __31__ for details.

## Issue 39

<span custom-style = 'reviewer'> You utilize the terms mental disease, mental stress, mental symptoms, etc… Be more specific and replace these terms by what you measured, i.e., HADS-anxiety, HADS-depression, and PSS. 
You mention somatic symptoms in the discussion but what they refer to remains obscure. 
You also refer to somatic diseases. 
Pathologists would rather speak of physical disorders as opposed to psychiatric disorders.</span>

We are grateful for this hint. 
In the revised manuscript we provide the psychometric scale when referencing depression, anxiety and stress signs. 
We removed the sentence of __Discussion__ referring to somatic symptoms, since those were not systematically surveyed in the SIMMUN cohort. 
Following your suggestion, we term 'somatic conditions' 'physical disorders' consequently in the revised text.

## Issue 40

<span custom-style = 'reviewer'>The discussion is convoluted and needs to be simplified. 
It would be easier to follow if you would first discuss the significance of the association between inflammation and neurotransmitter precursors and then the association between these two sets of factors and what you call mental health measured by HADS and PSS, first in the immune -> HDAS symptomatology or kynurenine -> HDAS symptomatology and then in the stress -> immune direction, remembering that in the case of your own cohort, the cross-sectional nature of your study does not allow you to get further than describing associations.</span>

We agree with the Reviewer that there was a lot of room for improvement in __Discussion__.
We rewrote this part of the manuscript during the revision according to the following scheme:

* A brief generalizing summary of the results obtained with both cohorts. In this initial paragraph, we highlight the most relevant factors for systemic serotonin and dopamine metabolism: inflammation, SARS-CoV-2 infection, TRP availability, age, signs of depression and mental stress. In this paragraph, we also discuss some discrepancy in our data such as way higher effects of inflammation on neurotransmitter availability in the INCOV cohort than in the SIMMUN dataset.

* Role of inflammation and SARS-CoV-2 in the KYN pathway and PHE - TYR conversion with implication of systemic availability of serotonin and dopamine. In this paragraph we discuss clinical literature and mechanistic reports tackling with IDO and PAH activity as well as BH4, the key co-factor of the serotonin and dopamine pathway enzymes.

* Effects of mental health-related factors (psychiatric disorders, mental stress) on systemic serotonin and dopamine availability.

* Relevance of systemic metabolism of serotonin and dopamine for the central nervous system and development of psychiatric disorders. Here we also underline that evidence for effects of peripheral serotonin, systemic activity of the KYN pathway and peripheral dopamine synthesis from PHE and TYR is limited and bases in many cases on clinical association of psychiatric disorders with blood levels of the neurotransmitters, their precursors or competitor pathway products.

## Issue 41

<span custom-style = 'reviewer'>Page 15: the study by Benedetti is at best suggestive but does not demonstrate anything.</span>

We agree with the Reviewer, that the report by Benedetti et al. does not provide any solid mechanistic link between the COVID-19-related inflammation and depression. 
We removed this paper from __Discussion__ of the revised manuscript.

# Reviewer 3

## Issue 42

<span custom-style = 'reviewer'>Considering the disease SARS CoV-2 and its consequences in the form of post-covid syndrome, the presented work is highly relevant.</span>

We thank the Reviewer for appreciation of our manuscript and the valuable feedback.

## Issue 43

<span custom-style = 'reviewer'>In the abstract, the authors introduce ...markers of serotonin availability (TRP, KYN, KYN/TRP ratio)....(line 24). 
This statement is not accurate, as Serotonin is formed from TRP, but there is no KYN in its catabolic pathway. 
Only if the catabolism of TRP moves more to the kynurenine pathway, less TRP goes to the serotonin pathway, and the availability of serotonin is reduced. 
This argument is similar to the line 44.</span>

We concur with the Reviewer's opinion. 
Certainly, kynurenine and the kynurenine/tryptophan ratio are markers of the pathway which competes with the serotonin biosynthesis. 
For the sake of clarity, we we term kynurenine, quinolinate and kynurenine/tryptophan ratio 'competitor pathway products/metabolites' or products/metabolites of the competitive kynurenine pathway' in the revised text.

## Issue 44

<span custom-style = 'reviewer'>The principle point is that it would help the reader to orient himself in such a large amount of data if the authors completed the manuscript with a design scheme of the INCOV and SIMMUN study</span>

To improve readability of the revised manuscript, we present the results in a cohort-wise manner, i.e. beginning with the SIMMUN cohort as an 'exploratory' collective and proceeding with the INCOV cohort as a king of 'validation' collective used to verify and study in more detail (e.g. by time course analyses) the key findings from the SIMMUN cohort. 
Please refer to __Issues 1__ and __Issue 38__ for details. 
The modified analysis strategy is presented in __Figure 1__.

## Issue 45

### 2023-05-23 paper/markdown/rebuttal_letter.Rmd

<span custom-style = 'reviewer'>In line 16 on p. 6 are introducing breakdown products as KYN, TRP, KYN/TRP ratio ..... It would be more precise to introduce them in the order they are formed: TRP, KYN, KYN/TRP ratio ....pls, correct everywhere in the manuscript similar p. 9, line 29 ..... For example, correct position is given on the line 52 of p.11</span>

Thanks for the hint. 
We stick in the revised manuscript to the proposed metabolite order.

## Issue 46

<span custom-style = 'reviewer'>p. 14, line 34 : a large meta-analyses ... include more citations</span>

We re-wrote the Discussion part of the revised manuscript and provide now more evidence [@Hunt2020; @Maes2012; @Brown2021; @Marx2020;  @Fellendorf2022] on the link between the so called sickness behavior, inflammation, serotonin and dopamine metabolism in the final part of the scion tackling the relevance of our findings for mental health. 
We also discuss the controversial relationship between circulating markers of neurotransmitter availability such as tryptophan or serotonin, and the competitor pathway products (kynurenine, quinolinate) on central nervous system and mental health [@Brown2021; @Moncrieff2022].

## Issue 47

<span custom-style = 'reviewer'>p. 15, line 38 - correct as a note in the abstract</span>

_Katharina, could you please respond? I don't have the submission text with the line numbering and can't find the fragment. Thanks!_

## Issue 48

<span custom-style = 'reviewer'>p. 15, line 62 - where ???</span>

_As above, thanks!_

## Issue 49

<span custom-style = 'reviewer'>put one line above the grey text stating that it is INCOR</span>

_Same as above!_

## Issue 50

<span custom-style = 'reviewer'> the given data do not match mathematically - check or explain why</span>

We apologize for the typos concerning numbers in the study inclusion scheme. 
Please refer to the correct numbers in the revised __Figure 1__.

## Issue 51

<span custom-style = 'reviewer'>explain used abbreviations in the text below</span>

We are sorry for that. 
We explain now all abbreviations used in __Figures__ in their legends.

\newpage

# References
